---
title: "STATS C216 Bayesian FP STEM Social Agency"
author: "Shuhan (Alice) Ai"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
    number-sections: true
execute:
  warning: false
  message: false
editor: visual
---

Models (all fit with `rstanarm`: `stan_glm` / `stan_glmer` / `stan_gamm4`). Set A uses default weakly informative priors; Set B replaces them with a regularized horseshoe on parametric coefficients (smooth-term priors unchanged).

- **Set A — default priors**
  - M1 Linear: y \~ comserv_freq + covariates
  - M2 Additive: y \~ comserv_freq + s(continuous...) + remaining covariates
  - M3 Mixed: y \~ comserv_freq + covariates + (1 \| UNITID)
  - M4 Additive mixed: y \~ comserv_freq + s(continuous...) + covariates + (1 \| UNITID)
- **Set B — sparsity (regularized horseshoe) priors**
  - M5 Linear + HS
  - M6 Additive + HS (smooths unshrunk)
  - M7 Mixed + HS
- **Set C — interactions under sparsity prior**
  - M8 Mixed + HS + comserv_freq × {gender, race, firstgen.b}

Outcome: `socagency.sd_smooth` (jittered to remove the IRT ceiling spike — see Data Preparation).

# Setup

```{r setup}
library(tidyverse)
library(rstanarm)    # Bayesian regression (course's main package)
library(loo)         # LOO / WAIC model comparison
library(bayesplot)   # posterior predictive checks, diagnostics
library(tidybayes)   # tidy posterior summaries
library(posterior)   # as_draws / summarise_draws for diagnostics
library(knitr)
library(dplyr)

# Reproducibility + use all cores for the 4 MCMC chains
options(mc.cores = parallel::detectCores())
set.seed(216)

# Sampler defaults for every fit below (bump iter if ESS/R-hat need it)
CHAINS <- 4
ITER   <- 2000      # 1000 warmup + 1000 sampling per chain by default
```

```{r load-data}
data <- readRDS("/Users/aishuhan/Desktop/PhD Dissertation/Data/HERICSS2021-25_STEMSRO_clean_v01.rds")
dim(data)
str(data)
```

# Data Preparation

## Smoothing the Ceiling Spike in Social Agency

The raw `socagency.sd` distribution shows a sharp ceiling spike near the top of the scale — an artifact of the IRT scoring procedure that places many students at a single maximum value. To reduce this discontinuity prior to Gaussian modelling, we add small Gaussian noise to all observations within $\varepsilon = 0.05$ of the maximum.

Specification: - Detect ceiling mask: $\{i : y_i \ge y_{\max} - \varepsilon\}$, with $\varepsilon = 0.05$ - Jitter: $y_i^{\text{smooth}} = y_i + \eta_i$, where $\eta_i \sim \mathcal{N}(0, 0.10^2)$ for $i$ in the ceiling mask; $y_i^{\text{smooth}} = y_i$ otherwise - Seeded for reproducibility (`set.seed(216)`)

This adds at most \~0.10 SD of noise to a small fraction of observations and leaves the bulk of the distribution untouched. All Bayesian models below use `socagency.sd_smooth` as the outcome.

**Reporting:** The ceiling spike is an IRT scoring artifact that violates the Gaussian likelihood assumption and would otherwise show up as spurious misspecification in posterior predictive checks. Light jitter on the ceiling values restores continuity in the latent construct that the IRT score is intended to measure.

```{r smooth-ceiling}
# 1. Inspect the spike
max_val <- max(data$socagency.sd, na.rm = TRUE)
eps     <- 0.05
ceiling_mask <- data$socagency.sd >= (max_val - eps)
n_ceiling    <- sum(ceiling_mask, na.rm = TRUE)
cat("Max value:", round(max_val, 3),
    "| Ceiling threshold (>=", round(max_val - eps, 3), "):",
    n_ceiling, "students (",
    round(100 * n_ceiling / nrow(data), 2), "% of sample )\n")

# 2. Jitter only the ceiling values
set.seed(216)
data$socagency.sd_smooth <- data$socagency.sd
data$socagency.sd_smooth[ceiling_mask] <-
  data$socagency.sd[ceiling_mask] + rnorm(n_ceiling, mean = 0, sd = 0.10)

# 3. Before / after comparison
par(mfrow = c(1, 2))
hist(data$socagency.sd,        breaks = 40,
     main = "Original socagency.sd", xlab = "socagency.sd")
hist(data$socagency.sd_smooth, breaks = 40,
     main = "Smoothed (jitter on ceiling)", xlab = "socagency.sd_smooth")
par(mfrow = c(1, 1))

# Summary stats: confirm minimal global distortion
summary(data$socagency.sd)
summary(data$socagency.sd_smooth)
```

# Descriptive Analysis

```{r}
library(table1)
descr <- table1(~ socagency.sd_smooth +
                  # school IVs
                  insttype + instcont + selectivity.sd + hbcu +
                  # student background (gender now 3 levels: Men/Women/GNC)
                  gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b, data = data)
descr
```

# Bayesian Models — Default (Weakly Informative) Priors

All four models below use `socagency.sd_smooth` as the outcome and `rstanarm` default priors (autoscaled, weakly informative Normal on coefficients; Exponential(1) on $\sigma$; `decov()` on the random-effect covariance in M3/M4).

## M1 — Linear (stan_glm)

```{r fit-m1-linear}
hist(data$socagency.sd_smooth, breaks = 40)

m1_linear <- stan_glm(
  socagency.sd_smooth ~
    # baseline social agency measures
    tfssocagency.sd + 
    # community service activity
    comserv_freq + tfsvolunteer +
    # school IVs
    insttype + instcont + selectivity.sd + hbcu +
    # student background (gender now 3 levels: Men/Women/GNC)
    gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b +
    # pre-college STEM
    ymath.sd + yphys.sd + ybio.sd + ycs.sd +
    # aspirations, values, abilities
    degreeasp.sub +
    tfshabits.sd + tfsacselfcon.sd + tfssocselfcon.sd,
  data   = data,
  family = gaussian(),
  chains = 4, iter = 2000, seed = 216, refresh = 0
)

# Coefficient table
summary(m1_linear, probs = c(0.025, 0.5, 0.975), digits = 3)
prior_summary(m1_linear)
```

## M2 — Additive (stan_gamm4)

```{r fit-m2-additive}
# M2 — Additive: selected continuous predictors enter as penalized splines
# s(); the rest stay linear. stan_gamm4() fits GAMs in rstanarm (it wraps mgcv's smooth construction).
m2_additive <- stan_gamm4(
  socagency.sd_smooth ~
    # baseline social agency measure
    s(tfssocagency.sd) +
    # community service activity
    comserv_freq + tfsvolunteer +
    # school IVs
    insttype + instcont + s(selectivity.sd) + hbcu +
    # student background
    gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b +
    # pre-college STEM
    ymath.sd + yphys.sd + ybio.sd + ycs.sd +
    # aspirations, values, abilities
    degreeasp.sub + s(tfshabits.sd) + s(tfsacselfcon.sd) + s(tfssocselfcon.sd),
  data = data, family = gaussian(),
  chains = CHAINS, iter = ITER, seed = 216, refresh = 0,
  adapt_delta = 0.99           # smooths sometimes need a higher target
)

# Coefficient table
summary(m2_additive, probs = c(0.025, 0.5, 0.975), digits = 3)
prior_summary(m2_additive)
```

## M3 — Mixed (stan_glmer)

```{r fit-m3-mixed}
# Calculate ICC first
library(lme4)

m0 <- lmer(socagency.sd_smooth ~ 1 + (1 | UNITID), data = data)
vc     <- as.data.frame(VarCorr(m0))
tau2   <- vc$vcov[vc$grp == "UNITID"]     # between-institution variance
sigma2 <- vc$vcov[vc$grp == "Residual"]   # within-institution (residual) variance
icc    <- tau2 / (tau2 + sigma2)
icc   # ICC = 0.044

# M3 — Mixed: same linear specification as M1 PLUS a random intercept by
# institution. stan_glmer() is rstanarm's lme4-style multilevel function;
# the group-variance prior is decov() by default.
m3_mixed <- stan_glmer(
  socagency.sd_smooth ~
    # baseline social agency measure
    tfssocagency.sd +
    # community service activity 
    comserv_freq + tfsvolunteer +
    # school IVs
    insttype + instcont + selectivity.sd + hbcu +
    # student background
    gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b +
    # pre-college STEM
    ymath.sd + yphys.sd + ybio.sd + ycs.sd +
    # aspirations, values, abilities
    degreeasp.sub + tfshabits.sd + tfsacselfcon.sd + tfssocselfcon.sd +
    # institution random intercept
    (1 | UNITID),
  data = data, family = gaussian(),
  chains = CHAINS, iter = ITER, seed = 216,
  adapt_delta = 0.95
)

# Coefficient table
summary(m3_mixed, probs = c(0.025, 0.5, 0.975), digits = 3)
prior_summary(m3_mixed)
```

## M4 — Additive mixed (stan_gamm4 + RI)

```{r fit-m4-addmixed}
# M4 — Additive mixed: the M2 smooths and a random intercept
m4_addmixed <- stan_gamm4(
  socagency.sd_smooth ~
    # baseline social agency measure
    s(tfssocagency.sd) +
    # community service activity (comserv_freq = treatment)
    comserv_freq + tfsvolunteer +
    # school IVs
    insttype + instcont + s(selectivity.sd) + hbcu +
    # student background
    gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b +
    # pre-college STEM
    ymath.sd + yphys.sd + ybio.sd + ycs.sd +
    # aspirations, values, abilities
    degreeasp.sub + s(tfshabits.sd) + s(tfsacselfcon.sd) + s(tfssocselfcon.sd),
  random = ~(1 | UNITID),
  data = data, family = gaussian(),
  chains = CHAINS, iter = ITER, seed = 216,
  adapt_delta = 0.99
)

# Coefficient table
summary(m4_addmixed, probs = c(0.025, 0.5, 0.975), digits = 3)
prior_summary(m4_addmixed)
```

# Bayesian Models — Sparsity (Regularized Horseshoe) Priors

## Horseshoe Specification

For the second set of models (M5–M7), we replace the default weakly informative Normal prior on each parametric coefficient with the **regularized horseshoe** (Piironen & Vehtari, 2017). The prior shrinks small effects toward zero while leaving large signals essentially unpenalized, which is well suited to a regression with \~30 covariates of mixed substantive relevance.

For coefficient $\beta_p$, $$\beta_p \mid \lambda_p, \tau, c \;\sim\; \mathcal{N}\!\left(0,\ \tau^2\, \tilde\lambda_p^{\,2}\right), \quad
\tilde\lambda_p^{\,2} = \frac{c^2 \lambda_p^{\,2}}{c^2 + \tau^2 \lambda_p^{\,2}},$$ with local scales $\lambda_p \sim \mathrm{Half\text{-}}t_{\nu}$, global scale $\tau \sim \mathrm{Half\text{-}}t_{\nu_g}(0, \tau_0)$, and slab scale $c^2 \sim \mathrm{Inv\text{-}Gamma}$. We set $\nu = \nu_g = 1$ (Cauchy local and global), slab df = 4, slab scale = 2.5, and the global scale $\tau_0$ following $\tau_0 \approx (p_0 / (p - p_0)) \cdot \sigma / \sqrt{n}$, with $p \approx 30$ parametric coefficients and an expected number of nonzero coefficients $p_0 \approx 5$. This gives $\tau_0 \approx 0.01$.

**Important:** in `stan_gamm4`, the `prior` argument applies only to the parametric (linear) coefficients. Smooth-term basis coefficients keep their own penalized priors, so applying the horseshoe leaves the additive smooths unshrunk — matching our design requirement.

```{r hs-spec}
# Global scale: tau0 = (p0 / (p - p0)) * sigma_y / sqrt(n)
# Use p = 30, p0 = 5; sigma_y on standardized outcome ~ 1
p_total   <- 30
p_nonzero <- 5
n_obs     <- nrow(data)
tau0      <- (p_nonzero / (p_total - p_nonzero)) / sqrt(n_obs)
tau0   # ~ 0.0026; round up to 0.01 for a slightly less aggressive global shrinkage

hs_prior <- hs(df = 1, global_df = 1, global_scale = 0.01,
               slab_df = 4, slab_scale = 2.5)
```

## M5 — Linear + horseshoe

```{r fit-m5-linear-hs}
m5_linear_hs <- stan_glm(
  socagency.sd_smooth ~
    # baseline social agency
    tfssocagency.sd +
    # community service activity
    comserv_freq + tfsvolunteer +
    # school IVs
    insttype + instcont + selectivity.sd + hbcu +
    # student background
    gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b +
    # pre-college STEM
    ymath.sd + yphys.sd + ybio.sd + ycs.sd +
    # aspirations, values, abilities
    degreeasp.sub + tfshabits.sd + tfsacselfcon.sd + tfssocselfcon.sd,
  data   = data,
  family = gaussian(),
  prior  = hs_prior,
  chains = CHAINS, iter = ITER, seed = 216, refresh = 0,
  adapt_delta = 0.99
)

summary(m5_linear_hs, probs = c(0.025, 0.5, 0.975), digits = 3)
prior_summary(m5_linear_hs)
```

## M6 — Additive + horseshoe (linear coefs only)

```{r fit-m6-additive-hs}
m6_additive_hs <- stan_gamm4(
  socagency.sd_smooth ~
    s(tfssocagency.sd) +
    comserv_freq + tfsvolunteer +
    insttype + instcont + s(selectivity.sd) + hbcu +
    gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b +
    ymath.sd + yphys.sd + ybio.sd + ycs.sd +
    degreeasp.sub + s(tfshabits.sd) + s(tfsacselfcon.sd) + s(tfssocselfcon.sd),
  data = data, family = gaussian(),
  prior  = hs_prior,           # applies to parametric coefs only; smooths unshrunk
  chains = CHAINS, iter = ITER, seed = 216, refresh = 0,
  adapt_delta = 0.99
)

summary(m6_additive_hs, probs = c(0.025, 0.5, 0.975), digits = 3)
prior_summary(m6_additive_hs)
```

## M7 — Mixed + horseshoe

```{r fit-m7-mixed-hs}
m7_mixed_hs <- stan_glmer(
  socagency.sd_smooth ~
    tfssocagency.sd +
    comserv_freq + tfsvolunteer +
    insttype + instcont + selectivity.sd + hbcu +
    gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b +
    ymath.sd + yphys.sd + ybio.sd + ycs.sd +
    degreeasp.sub + tfshabits.sd + tfsacselfcon.sd + tfssocselfcon.sd +
    (1 | UNITID),
  data = data, family = gaussian(),
  prior  = hs_prior,           # fixed-effect coefs; RI keeps decov()
  QR     = TRUE,               # orthogonalize design matrix (faster sampling)
  chains = CHAINS, iter = ITER, seed = 216, refresh = 0,
  adapt_delta = 0.99
)

summary(m7_mixed_hs, probs = c(0.025, 0.5, 0.975), digits = 3)
prior_summary(m7_mixed_hs)
```

# Bayesian Models — Interactions under Sparsity Prior

## Rationale

With the regularized horseshoe absorbing uninformative terms, we can safely introduce interaction blocks to probe heterogeneity in the community-service association without inflating posterior uncertainty for the main effects. Drawing on the dissertation finding that women, students of color, and first-generation STEM students show higher baseline social responsibility, we test whether the community-service association *differs* by these three demographic dimensions:

- `comserv_freq × gender`
- `comserv_freq × race`
- `comserv_freq × firstgen.b`

We do not include institutional interactions (HBCU, major) here, keeping the model focused on the demographic heterogeneity claims most aligned with the dissertation story.

## M8 — Mixed + horseshoe + demographic interactions

```{r fit-m8-mixed-hs-int}
m8_mixed_hs_int <- stan_glmer(
  socagency.sd_smooth ~
    tfssocagency.sd +
    comserv_freq + tfsvolunteer +
    insttype + instcont + selectivity.sd + hbcu +
    gender + race + tfsmajor.stemsub + firstgen.b + hsgpa.sd + parentstem.b +
    ymath.sd + yphys.sd + ybio.sd + ycs.sd +
    degreeasp.sub + tfshabits.sd + tfsacselfcon.sd + tfssocselfcon.sd +
    # demographic interactions with treatment
    comserv_freq:gender + comserv_freq:race + comserv_freq:firstgen.b +
    (1 | UNITID),
  data = data, family = gaussian(),
  prior  = hs_prior,
  QR     = TRUE,               # orthogonalize design matrix (faster sampling)
  chains = CHAINS, iter = ITER, seed = 216, refresh = 0,
  adapt_delta = 0.99
)

summary(m8_mixed_hs_int, probs = c(0.025, 0.5, 0.975), digits = 3)

# Pull out only the interaction coefficients for inspection
int_pars <- grep("^comserv_freq.*:", rownames(summary(m8_mixed_hs_int)), value = TRUE)
summary(m8_mixed_hs_int, pars = int_pars,
        probs = c(0.025, 0.5, 0.975), digits = 3)
```

# Convergence and Model Checking

```{r diagnostics}
fits <- list(
  `M1 Linear`           = m1_linear,
  `M2 Additive`         = m2_additive,
  `M3 Mixed`            = m3_mixed,
  `M4 Additive mixed`   = m4_addmixed,
  `M5 Linear + HS`      = m5_linear_hs,
  `M6 Additive + HS`    = m6_additive_hs,
  `M7 Mixed + HS`       = m7_mixed_hs,
  `M8 Mixed + HS + Int` = m8_mixed_hs_int
)

# R-hat and bulk-ESS check for every model (want R-hat ~ 1.00, ESS in 1000s)
# as.matrix() on a stanreg object returns the posterior draws matrix.
purrr::imap_dfr(fits, function(m, nm) {
  s <- posterior::summarise_draws(as.matrix(m), "rhat", "ess_bulk")
  tibble(model = nm,
         max_rhat = max(s$rhat, na.rm = TRUE),
         min_ess  = min(s$ess_bulk, na.rm = TRUE))
}) %>% kable(digits = 3, caption = "Convergence summary across models")
```

```{r ppcheck}
# Posterior predictive checks — does each model reproduce the outcome dist
pp_check(m1_linear,      ndraws = 100) + ggtitle("M1 Linear")
pp_check(m2_additive,    ndraws = 100) + ggtitle("M2 Additive")
pp_check(m3_mixed,       ndraws = 100) + ggtitle("M3 Mixed")
pp_check(m4_addmixed,    ndraws = 100) + ggtitle("M4 Additive mixed")
pp_check(m5_linear_hs,   ndraws = 100) + ggtitle("M5 Linear + HS")
pp_check(m6_additive_hs, ndraws = 100) + ggtitle("M6 Additive + HS")
pp_check(m7_mixed_hs,    ndraws = 100) + ggtitle("M7 Mixed + HS")
pp_check(m8_mixed_hs_int,ndraws = 100) + ggtitle("M8 Mixed + HS + Int")
```

```{r smooth-plots}
# Visualize the estimated 
# plot_nonlinear() is rstanarm's helper for stan_gamm4 smooth term
plot_nonlinear(m2_additive)
```

# Model Comparison (Robustness)

```{r loo-compare}
loo_m1 <- loo(m1_linear)
loo_m2 <- loo(m2_additive)
loo_m3 <- loo(m3_mixed)
loo_m4 <- loo(m4_addmixed)        # overfitting under default priors
loo_m5 <- loo(m5_linear_hs)
loo_m6 <- loo(m6_additive_hs)
loo_m7 <- loo(m7_mixed_hs)
loo_m8 <- loo(m8_mixed_hs_int)

# loo_compare orders models best-first
loo_compare(loo_m1, loo_m2, loo_m3, loo_m4,
            loo_m5, loo_m6, loo_m7, loo_m8)

print(loo_m4)
print(loo_m8)
```

# Treatment Effect Summary

## Posterior of `comserv_freq` across all models

```{r tx-effect-summary}
# Pull posterior draws for the main treatment coefficient from every fit
tx_par <- "comserv_freqTreated"

tx_draws <- purrr::imap_dfr(fits, function(m, nm) {
  d <- as.data.frame(m, pars = tx_par)
  tibble(model = nm, beta = d[[tx_par]])
})

# Posterior summary table
tx_summary <- tx_draws %>%
  group_by(model) %>%
  summarise(
    mean   = mean(beta),
    sd     = sd(beta),
    q025   = quantile(beta, 0.025),
    q975   = quantile(beta, 0.975),
    p_gt_0 = mean(beta > 0)
  )
tx_summary %>% kable(digits = 3,
  caption = "Posterior of community-service association across models")

# Posterior density overlay
ggplot(tx_draws, aes(x = beta, colour = model)) +
  geom_density(linewidth = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = expression(beta[comserv_freq]),
       y = "Posterior density",
       title = "Posterior of community-service association across models") +
  theme_minimal()
```

## M8 — demographic interaction posteriors

```{r m8-interactions}
# Extract interaction-coefficient posteriors and plot 95% CrIs
all_pars <- rownames(summary(m8_mixed_hs_int))
int_pars <- grep("^comserv_freq.*:|^comserv_freqTreated:", all_pars, value = TRUE)
int_pars

int_draws <- as.data.frame(m8_mixed_hs_int, pars = int_pars) %>%
  tidyr::pivot_longer(everything(), names_to = "term", values_to = "beta")

int_draws %>%
  group_by(term) %>%
  summarise(
    mean = mean(beta), sd = sd(beta),
    q025 = quantile(beta, 0.025), q975 = quantile(beta, 0.975),
    p_gt_0 = mean(beta > 0)
  ) %>%
  arrange(desc(abs(mean))) %>%
  kable(digits = 3,
        caption = "M8: posterior of comserv × demographic interaction terms")

ggplot(int_draws, aes(x = beta, y = term)) +
  tidybayes::stat_halfeye(.width = c(0.5, 0.95)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Interaction coefficient (on standardized outcome)",
       y = NULL,
       title = "M8: heterogeneity of community-service association") +
  theme_minimal()
```
